Learning constraints in spreadsheets and tabular data

### 2016-05-28 Paper/ECAI/tabular.tex
Rewrite table definitions

  \begin{subfigure}{.65\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicate headerless tables, dark borders indicate maximal type-consistent blocks.
    This example was created by combining several smaller Excel sheets originating from an exercise session.
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.35\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}}\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^1, \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^1 \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^1 , \ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^1 \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^1 \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\eclookup{\range{T_{4}}{\rangeall}{2}}{\range{T_{4}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^2 \\
%
      &~\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
	  % Table 6
      &~\ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}}
  \vspace{-10pt}
  \caption{Constraints present in the running example (left), except \textit{ALLDIFFERENT} (16), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).
  $^1$(Spurious constraints) $^2$(Not present in original source)}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

\begin{table*}
\caption{Overview of constraint templates implemented in \sname.
The templates marked with $\dagger$ are aggregate constraint templates, the table shows the version for sum but the implementation also supports max, min, average, product and count.
Subgroups in normal font have to be vectors, while subgroups in \textbf{bold} may contain more than one vector.
Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.}\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sg_x}^*$
      & $\discrete(\sg_x)$
      & All values in $\sg_x$ are different: $\sg_x[i] \neq \sg_x[j]$ if $i \neq j$
      \\ \hline
    %X = Y & & \\
    $\ecperm{\sg_x}^*$
      & $\numeric(\sg_{x})$, $\ecalldiff{\sg_{x}}$
      & The values in $\sg_{x}$ are a permutation of the numbers $1$ through $\plength(\sg_{x})$.
      \\ \hline
    \ecseries{\sg_x}
      & $\integer(\sg_{x})$ and $\ecperm{\sg_{x}}$
      & $\sg_{x}[1] = 1$ and $\sg_{x}[i] = \sg_{x}[i - 1] + 1$.
      \\ \hline
    $\ecfkey{\sg_{fk}}{\sg_{pk}}^*$ & $\sg_{fk}$ and~$\sg_{pk}$ are both $\discrete$; $\ptype(\sg_{fk}) = \ptype(\sg_{pk})$; $\ptable(\sg_{fk}) \neq \ptable(\sg_{pk})$; and $\ecalldiff{\sg_{pk}}$ & Every value in~$\sg_{fk}$ also exist in~$\sg_{pk}$ \\ \hline
    % TODO check out
    \eclookup{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & $\sg_{fk}$ and $\sg_{pk}$ are both $\discrete$; arguments $\{\sg_{fk}, \sg_{r}\}$ and $\{\sg_{pk}, \sg_{val}\}$ within the same set have the same \plength, \ptable and \por; $\sg_{r}$ and~$\sg_{val}$ have the same type; and \ecfkey{\sg_{fk}}{\sg_{pk}}.
      & $\sg_r[i]$ is the same value as looking up~$\sg_{fk}[i]$ in~$\sg_{pk}$  and returning the corresponding value in~$\sg_{val}$.
      \\ \hline
    \eclookupfuzzy{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & \textit{Same as lookup}
      & $\sg_r[i]$ is the same value as looking up the last item in~$\sg_{pk}$ smaller than~$\sg_{fk}[i]$ and returning the corresponding value in~$\sg_{val}$.
      \\ \hline
    \eclookupprod{\sg_r}{\sg_1}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{fk}\}$ are $\numeric$, arguments $\{\sg_{pk}, \sg_{val}\}$ are $\discrete$ and within both sets all arguments have the same \plength, \ptable and \por; also \ecfkey{\sg_{fk}}{\sg_{pk}}.
      & $\sg_{r}[i]$ is the obtained by multiplying $\sg_{1}[i]$ with $\eclookupf{\sg_{fk}}{\sg_{pk}}{\sg_{val}}{}[i]$.
      \\ \hline
    \ecprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$
      & $\sg_{r}[i] = \sg_{1}[i] \times \sg_{2}[i]$.
      \\ \hline
    \ecdiff{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$ and $ \por$
      & $\sg_{r}[i] = \sg_{1}[i] - \sg_{2}[i]$.
      \\ \hline
    \ecproj{\sg_r}{\mathbf{\sg_x}}
      & Arguments $\{\sg_{r}, \mathbf{\sg_x}\}$ all have the same $\plength$, $\por$, $\ptable$ and $\ptype$; $\mathbf{\sg_x}$ contains at least~2 vectors; and $\sg_r = \mathit{SUM}(\mathbf{\sg_x}, \por(\mathbf{\sg_x}))$
      & At every position~$i$ in $1$ through $\plength(\sg_{r})$ there is exactly one vector~$v$ in $\mathbf{\sg_x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sg_{r}[i]$.
      \\ \hline
    \ecrank{\sg_r}{\sg_x}
      & $\integer(\sg_{r})$; $\numeric(\sg_{x})$; and $\plength(\sg_{r}) = \plength(\sg_{x})$
      & The values in $\sg_{r}$ represent the rank (from largest to smallest) of the values in $\sg_{x}$ (including ties)
      \\ \hline
    \ectotal{\sg_r}{\sg_{pos}}{\sg_{neg}}
      & Arguments $\{\sg_{r}, \sg_{pos}, \sg_{neg}\}$ are all $\numeric$ and all have the same $\plength$, which is at least $2$
      & The values in $\sg_{r}$ are a running total, each value $\sg_{r}[i] = \sg_{r}[i - 1] + \sg_{pos}[i] - \sg_{neg}[i]$.
      \\ \hline
    $\ecsumr{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\pcols(\mathbf{\sg_x}) \geq 2$; and $\prows(\mathbf{\sg_x}) = \plength(\sg_r)$
      & Each value in $\sg_{r}$ is obtained by summing over the corresponding row in $\mathbf{\sg_x}$.
      \\ \hline
    $\ecsumc{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\prows(\mathbf{\sg_x}) \geq 2$; and $\pcols(\mathbf{\sg_x}) = \plength(\sg_r)$
      & Each value in $\sg_{r}$ is obtained by summing over the corresponding column in $\mathbf{\sg_x}$.
      \\ \hline
    $\ecsumif{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}^\dagger$
      & $\sg_{fk}, \sg_{pk}$ are $\discrete$; $\sg_{r}, \sg_{val}$ are $\numeric$; within the sets $\{\sg_{val}, \sg_{fk}\}$ and $\{\sg_{pk}, \sg_{r}\}$ arguments have the same $\plength$ and $\por$; $\sg_{fk}$ and $\sg_{val}$ have the same $\ptable$; $\sg_{fk}$ and $\sg_{pk}$ must have different $\ptable$s but the same $\ptype$; and \ecalldiff{\sg_{pk}}
      & The value for $\sg_{r}[i]$ is obtained by summing all values $\sg_{val}[j]$ where $\sg_{fk}[j] = \sg_{pk}[i]$
      \\ \hline
    \ecsumprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_r, \sg_1, \sg_2\}$ are all $\numeric$; $\plength(\sg_{1}) = \plength(\sg_{2}) \geq 2$; and $\prows(\sg_{r}) = \pcols(\sg_{r}) = 1$
      & $\sg_{r}[i] = \sum_{i = 1}^{\plength(\sg_{1})} \sg_{1}[i] \times \sg_{2}[i]$.
  \end{tabularx}}

>>>>

  \begin{subfigure}{.65\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicate headerless tables, dark borders indicate maximal type-consistent blocks.
    Most tables only contain type-consistent columns, however, the data in table~$T_2$ can be used as both column- and row vectors.
    This example was created by combining several smaller Excel sheets originating from an exercise session.
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.35\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}}\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^1, \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^1 \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^1 , \ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^1 \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^1 \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\eclookup{\range{T_{4}}{\rangeall}{2}}{\range{T_{4}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^2 \\
%
      &~\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
	  % Table 6
      &~\ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}}
  \vspace{-10pt}
  \caption{Constraints present in the running example (left), except \textit{ALLDIFFERENT} (16), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).
  $^1$(Spurious constraints) $^2$(Not present in original source)}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

\begin{table*}
\caption{
  Overview of constraint templates implemented in \sname.
  Templates marked with $\dagger$ are \textit{aggregate} templates, the table shows the version for sum but \sname also supports max, min, average, product and count.
  Subgroups in normal font have to be vectors, whereas subgroups in \textbf{bold} may contain more than one vector.
  Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.
  $\discrete(x)$ is a shortcut for: $\textual(x) \lor \numeric(x)$.
}\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sg_x}^*$
      & $\discrete(\sg_x)$

      & All values in $\sg_x$ are different: $\sg_x[i] \neq \sg_x[j]$ if $i \neq j$
      \\ \hline
    %X = Y & & \\

    $\ecperm{\sg_x}^*$
      & $\numeric(\sg_{x})$, $\ecalldiff{\sg_{x}}$

      & The values in $\sg_{x}$ are a permutation of the numbers $1$ through $\plength(\sg_{x})$.
      \\ \hline

    \ecseries{\sg_x}
      & $\integer(\sg_{x})$ and $\ecperm{\sg_{x}}$

      & $\sg_{x}[1] = 1$ and $\sg_{x}[i] = \sg_{x}[i - 1] + 1$.
      \\ \hline

    $\ecfkey{\sg_{fk}}{\sg_{pk}}^*$
      & $\sg_{fk}$ and~$\sg_{pk}$ are both $\discrete$; $\ptype(\sg_{fk}) = \ptype(\sg_{pk})$; $\ptable(\sg_{fk}) \neq \ptable(\sg_{pk})$; and $\ecalldiff{\sg_{pk}}$

      & Every value in~$\sg_{fk}$ also exist in~$\sg_{pk}$ \\ \hline

    \eclookup{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & $\sg_{fk}$ and $\sg_{pk}$ are both $\discrete$; arguments $\{\sg_{fk}, \sg_{r}\}$ and $\{\sg_{pk}, \sg_{val}\}$ within the same set have the same \plength, \ptable and \por; $\sg_{r}$ and~$\sg_{val}$ have the same type; and \ecfkey{\sg_{fk}}{\sg_{pk}}.

      & $\sg_r[i]$ is the same value as looking up~$\sg_{fk}[i]$ in~$\sg_{pk}$  and returning the corresponding value in~$\sg_{val}$.
      \\ \hline

    \eclookupfuzzy{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & \textit{Same as lookup}

      & $\sg_r[i]$ is the same value as looking up the last item in~$\sg_{pk}$ smaller than~$\sg_{fk}[i]$ and returning the corresponding value in~$\sg_{val}$.
      \\ \hline

    \eclookupprod{\sg_r}{\sg_1}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{fk}\}$ are $\numeric$, arguments $\{\sg_{pk}, \sg_{val}\}$ are $\discrete$ and within both sets all arguments have the same \plength, \ptable and \por; also \ecfkey{\sg_{fk}}{\sg_{pk}}.

      & $\sg_{r}[i]$ is the obtained by multiplying $\sg_{1}[i]$ with $\eclookupf{\sg_{fk}}{\sg_{pk}}{\sg_{val}}{}[i]$.
      \\ \hline

    \ecprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$

      & $\sg_{r}[i] = \sg_{1}[i] \times \sg_{2}[i]$.
      \\ \hline

    \ecdiff{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$ and $ \por$

      & $\sg_{r}[i] = \sg_{1}[i] - \sg_{2}[i]$.
      \\ \hline

    \ecproj{\sg_r}{\mathbf{\sg_x}}
      & Arguments $\{\sg_{r}, \mathbf{\sg_x}\}$ all have the same $\plength$, $\por$, $\ptable$ and $\ptype$; $\mathbf{\sg_x}$ contains at least~2 vectors; and $\sg_r = \mathit{SUM}_{\por(\mathbf{\sg_x})}(\mathbf{\sg_x})$

      & At every position~$i$ in $1$ through $\plength(\sg_{r})$ there is exactly one vector~$v$ in $\mathbf{\sg_x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sg_{r}[i]$.
      \\ \hline

    \ecrank{\sg_r}{\sg_x}
      & $\integer(\sg_{r})$; $\numeric(\sg_{x})$; and $\plength(\sg_{r}) = \plength(\sg_{x})$

      & The values in $\sg_{r}$ represent the rank (from largest to smallest) of the values in $\sg_{x}$ (including ties)
      \\ \hline

    \ectotal{\sg_r}{\sg_{pos}}{\sg_{neg}}
      & Arguments $\{\sg_{r}, \sg_{pos}, \sg_{neg}\}$ are all $\numeric$ and all have the same $\plength$, which is at least $2$

      & $\sg_{r}[i] = \sg_{r}[i - 1] + \sg_{pos}[i] - \sg_{neg}[i]$.
      \\ \hline

    $\ecsumr{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\pcols(\mathbf{\sg_x}) \geq 2$; and $\prows(\mathbf{\sg_x}) = \plength(\sg_r)$

      & $B_r[i] = \sum ($row~$i$ of $\mathbf{B_x})$
      \\ \hline

    $\ecsumc{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\prows(\mathbf{\sg_x}) \geq 2$; and $\pcols(\mathbf{\sg_x}) = \plength(\sg_r)$

      & $B_r[i] = \sum ($column~$i$ of $\mathbf{B_x})$
      \\ \hline

    $\ecsumif{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}^\dagger$
      & $\sg_{fk}, \sg_{pk}$ are $\discrete$; $\sg_{r}, \sg_{val}$ are $\numeric$; within the sets $\{\sg_{val}, \sg_{fk}\}$ and $\{\sg_{pk}, \sg_{r}\}$ arguments have the same $\plength$ and $\por$; $\sg_{fk}$ and $\sg_{val}$ have the same $\ptable$; $\sg_{fk}$ and $\sg_{pk}$ must have different $\ptable$s but the same $\ptype$; and \ecalldiff{\sg_{pk}}

      & 
      \[
        B_r[i] = \sum_{j=1}^{\plength(B_{\textit{val}})} \begin{cases}
          \sg_{val}[j] & \text{if } \sg_{fk}[j] = \sg_{pk}[i] \\
          0 & \text{otherwise}
        \end{cases}
      \] \\ \hline

    \ecsumprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_r, \sg_1, \sg_2\}$ are all $\numeric$; $\plength(\sg_{1}) = \plength(\sg_{2}) \geq 2$; and $\prows(\sg_{r}) = \pcols(\sg_{r}) = 1$

      & $\sg_{r}[1] = \sum_{j = 1}^{\plength(\sg_{1})} \sg_{1}[j] \times \sg_{2}[j]$.
  \end{tabularx}}
